Predicting the (non-cryptographic) PCG Pseudo-Random Number Generator In Practice

### 2020-06-01 article/main.tex
fin section 5

Let us set:
\begin{equation}\label{eq:nabla}
  \nabla S_i \stackrel{def}{=} S_i - S_0 = \sum_{j=1}^i S_j - S_{j-1} \equiv \sum_{j=0}^{i-1} \Delta S_j \equiv \Delta S_0 \cdot \sum_{j=0}^{i-1} a^j \equiv \Delta S_0 \frac{a^i-1}{a-1} \mod 2^{128}
\end{equation}
Note that $\nabla S_0 = 0$ and $\nabla S_1 = \Delta S_0$. Therefore, knowledge
of $\Delta_0$ entails that of the whole sequence of $\nabla S_i$. The prediction
algorithm we propose proceeds in three phases:
\begin{enumerate}
\item Reconstruct $\Delta S_0[0:64+\ell]$ from $X_0, \dots, X_{4}$, check consistency with $X_5, \dots, X_{63}$.
\item Reconstruct all rotations $r_i$ from this partial knowledge.
\item Fully reconstruct $\Delta S_0$ from the rotations.
\item Reconstruct $S_0$ from $\Delta S_0$ and the rotations.
\end{enumerate}

@@

\Procedure{ReconstructPartialDifference}{$X_0, \dots, X_k$}
  \State // \emph{Statement involving $j$ must be repeated for $j=0, 1, 2, 3, 4$.}
  \State $H \gets$ LLL reduction of $G_{4,64}$
  \State $\ell \gets 14$
\For{$0 \leq w_0 < 2^{\ell}$ and $0 \leq c_0 < 2^{\ell - 1}$} \Comment{Guess least-significant bits}
\State $K_j \gets a^j w_0 + (2c_0 + 1)(a^j - 1)(a-1)^{-1} \bmod 2^{128}$ \Comment{Known part}
  \For{$0 \leq r_0, r_1, r_2, r_3, r_4 < 64$} \Comment{Guess rotations}
  \State $Y_j \gets X_j \lll r_j$ \Comment{Undo rotations}
  \State $T_j \gets \left(r_j \oplus Y_j[58:64]\right) +  64 \cdot \left(K_j \oplus Y_j\right)[0:\ell]$ \Comment{Truncated LCG}
  \State $\widetilde{T}'_j \gets T_j \boxminus  K_i[58:64+\ell]$ \Comment{Cancel known part}
  \State $\widetilde{\Delta T'}_j = \widetilde{T'}_{j+1} \boxminus \widetilde{T}'_j$  \Comment{Difference (truncated geom. seq.)}
  \State $(\Delta U_0, \dots, \Delta U_3) \gets \left\lfloor (\Delta T'_0, \dots, \Delta T'_3) \cdot 2^{58-\ell} \cdot \widetilde H^{-1} \right\rceil \cdot \widetilde H$ \Comment{CVP}
  \State $\Delta S_0[0:64+\ell] \gets \left(K_1 \boxminus K_0\right)[0:\ell]
  + 2^{\ell} \cdot \Delta U_0[0:64]$ \Comment{Check}
  \If{$\Call{ConsistencyCheck}{\Delta_0, w_0, Y_0, X_5, \dots, X_k}$}
  \State \Return $(w_0, c, r_0, \dots, r_4, \Delta_0)$.
  \EndIf
  \EndFor
  \EndFor
  \EndProcedure
\end{algorithmic}
\caption{Partial difference reconstruction algorithm (when $c$ is unknown).}
\label{algo:unknown_1}
\end{algorithm}

@@

The whole procedure is shown in algorithm~\ref{algo:last}.

@@

We have implemented the state reconstruction algorithms described above using a
mixture of \textsf{C} (for the computationally expensive parts) and
\textsf{Python} (for the rest). We used the \texttt{fplll} library~\cite{fplll}
to solve CVP instances exactly in dimension 42.

In this section, we briefly outline important aspects of our implementations and
present practical results. Our codes are available online at:
\[
  % \url{https://github.com/cbouilla/pcg/}
  \verb|[----REDACTED FOR PEER REVIEW----]|
\]
(and in the supplementary material).

The designer of \textsf{PCG} was kind enough to send us two sets challenge
inputs: one with the default (known) increment and one with a random secret
increment. She generated random seeds and provided us with the first outputs of
the pseudo-random generator. We were able to reconstruct the seed with an
extremely high confidence level, because they re-generate the same outputs. We
emailed back the seeds and received confirmation that they were indeed correct.

>>>>

Let us set:
\begin{equation}\label{eq:nabla}
  \nabla S_i \stackrel{def}{=} S_i - S_0 \equiv \sum_{j=0}^{i-1} \Delta S_j \equiv \Delta S_0 \cdot \sum_{j=0}^{i-1} a^j \equiv \Delta S_0 \frac{a^i-1}{a-1} \mod 2^{128}
\end{equation}
Note that $\nabla S_0 = 0$ and $\nabla S_1 = \Delta S_0$. Therefore, knowledge
of $\Delta_0$ entails that of the whole sequence of $\nabla S_i$. The prediction
algorithm we propose proceeds in three phases:
\begin{enumerate}
\item Reconstruct $\Delta S_0[0:64+\ell]$ from $X_0, \dots, X_{4}$, check consistency with $X_5, \dots, X_{63}$.
\item Reconstruct all rotations $r_i$ from this partial knowledge.
\item Fully reconstruct $\Delta S_0$ from the rotations.
\item Reconstruct $S_0$ from $\Delta S_0$ and the rotations.
\end{enumerate}

@@

\Procedure{ReconstructPartialDifference}{$X_0, \dots, X_k$}
  \State // \emph{Statement involving $j$ must be repeated for $j=0, 1, 2, 3, 4$.}
  \State $H \gets$ LLL reduction of $G_{4,64}$
  \State $\ell \gets 14$
\For{$0 \leq w_0 < 2^{\ell}$ and $0 \leq c_0 < 2^{\ell - 1}$} \Comment{Guess least-significant bits}
\State $K_j \gets a^j w_0 + (2c_0 + 1)(a^j - 1)(a-1)^{-1} \bmod 2^{128}$ \Comment{Known part}
  \For{$0 \leq r_0, r_1, r_2, r_3, r_4 < 64$} \Comment{Guess rotations}
  \State $Y_j \gets X_j \lll r_j$ \Comment{Undo rotations}
  \State $T_j \gets \left(r_j \oplus Y_j[58:64]\right) +  64 \cdot \left(K_j \oplus Y_j\right)[0:\ell]$ \Comment{Truncated LCG}
  \State $\widetilde{T}'_j \gets T_j \boxminus  K_i[58:64+\ell]$ \Comment{Cancel known part}
  \State $\Delta \widetilde{T}'_j = \widetilde{T}'_{j+1} \boxminus \widetilde{T}'_j$  \Comment{Difference (truncated geom. seq.)}
  \State $(\Delta U_0, \dots, \Delta U_3) \gets \left\lfloor (\Delta \widetilde{T}'_0, \dots, \Delta \widetilde{T}'_3) \cdot 2^{58-\ell} \cdot \widetilde H^{-1} \right\rceil \cdot \widetilde H$ \Comment{CVP}
  \State $\Delta S_0[0:64+\ell] \gets \left(K_1 \boxminus K_0\right)[0:\ell]
  + 2^{\ell} \cdot \Delta U_0[0:64]$ \Comment{Check}
  \If{$\Call{ConsistencyCheck}{\Delta_0, w_0, Y_0, X_5, \dots, X_k}$}
  \State \Return $(w_0, c_0, r_0, \dots, r_4, \Delta S_0)$.
  \EndIf
  \EndFor
  \EndFor
  \EndProcedure
\end{algorithmic}
\caption{Partial difference reconstruction algorithm (when $c$ is unknown).}
\label{algo:unknown_1}
\end{algorithm}

@@

The whole procedure is shown in algorithm~\ref{algo:last}. Note that once $S_0$
has been found, then all subsequent states can be computed with error using
$S_i = S_0 \boxplus \nabla S_i$. In particular, computing $S_1$ gives $c$ by
$c \gets S_1 \boxminus a S_0$. This complete the reconstruction procedure for
\textsf{PCG64}.

@@

We have implemented the state reconstruction algorithms described above using a
mixture of \textsf{C} (for the computationally expensive parts) and
\textsf{Python} (for the rest). We used the \texttt{fplll} library~\cite{fplll}
to solve CVP instances exactly in dimension 63.

In this section, we briefly outline important aspects of our implementations and
present practical results. Our codes are available in the supplementary material
as well as online at:
\[
  % \url{https://github.com/cbouilla/pcg/}
  \verb|[----REDACTED FOR PEER REVIEW----]|
\]
The designer of \textsf{PCG} was kind enough to send us two sets challenge
inputs: one with the default (known) increment and one with a random secret
increment. She generated random seeds and provided us with the first outputs of
the pseudo-random generator. We were able to reconstruct the seed with an
extremely high confidence level, because they re-generate the same outputs. We
emailed back the seeds and received confirmation that they were indeed correct.
